---
title: "A Report on Lake Ontario's Microbes"
author: "Kalem Hanlon"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# make document environment
```{r packages}
#put our code here
# load packages
library(tidyverse)
```

# load in the Lake Ontario data
```{r load data}
#load in lake ontario microbial community data
sample_and_taxon <- 
  read_csv("data/sample_and_taxon.csv")

# inspect the data
glimpse(sample_and_taxon)
```

#Microbial Cell Abundance Versus Temperature

```{r cells_vs_temp}
ggplot(data = sample_and_taxon, aes(x = temperature, y = cells_per_ml/1000000)) +
  labs(x = "Temperature (C)", y = "Cell Abundance (Millions)", title = "Microbial Cell Abundance in Lake Ontario by Temperature") +
  geom_point(aes(colour = env_group)) +
  theme_classic() +
  theme(legend.position = "bottom", legend.title = element_blank()) +
  geom_smooth(method = "lm", colour = "black")
```

## Cell abundance

The cell abundance tends to increase with the temperature. There appears to be a break in the data where there was no sampling between May and September.
- Deep samples are the coldest and have the fewest abundance of cells

```{r data summary, echo = FALSE}
#how many samples do I have
n_samples <- 
  sample_and_taxon |>
  #only pick out sample ID column
  select(sample_id) |> 
  #number of distinct names in the column
  n_distinct()
  
#calculate the minimum and max temps
#min
min_temp <- sample_and_taxon |> 
  summarise(round(min(temperature), digits = 2)) |> 
  pull()
#max
max_temp <- sample_and_taxon |> 
  summarise(round(max(temperature), digits = 2)) |> 
  pull()
```

The total number of samples is `r n_samples`. For this set of samples, temperature ranges from a minimum of `r min_temp`, and a maximum of `r max_temp` celsius.